---
title: "Motor Trend: Estimating MPG"
author: "Hugo Janssen"
date: "July 23, 2015"
output: pdf_document
---



```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, echo = TRUE)
options(xtable.comment = FALSE)
options(digits = 2)
```

```{r, echo = FALSE}
library(xtable)
library(ggplot2)
data(mtcars)
```



This edition of Motor Trend magazine explores the relationship between a set of variables and miles per gallon (MPG) for a collection of cars. In particular the following two questions are addressed:

* Is an automatic or manual transmission better for MPG?
* How to quantify the MPG difference between automatic and manual transmissions?

Source: `mtcars {datasets}` the data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).



# Executive Summary

The article starts with an exploratory analysis in which a baseline model is established based on transmission type alone. It is found that this model explains only 36% of the variance. In the data analysis that follows, more variables are introduced in the model. The proposed formula for the best fit is `mpg ~ 9.62 + 2.94 * am - 3.92 * wt + 1.23 * qsec` which explains 83% of the variance with a 95% confidence.



# Data Overview

The data set consists of a data frame with 32 observations on 11 variables. The first 4 entries of the data set are shown in table 1.

```{r, results = 'asis', echo = FALSE}
xtable(head(mtcars, 4), caption = "The first entries of the mtcars table", digits=c(0,1,0,0,0,2,2,2,0,0,0,0))
```



# Exploratory Analysis

The first model to fit is an univariate model where MPG is predicted by the transmission type alone. This provides a benchmark against which other models can be evaluated. The resulting coefficients are listed in table 2.

```{r}
fit0 <- lm(mpg ~ am, data = mtcars)
```

```{r, results = 'asis', echo = FALSE}
xtable(summary(fit0)$coefficients, caption = "Coefficients of a univariate model considering transmission type.", digits=c(0,2,2,2,6))
```

This model estimates an expected `r summary(fit0)$coefficients[2,1]` increase in MPG for cars with manual transmission compared to cars with automatic transmission. 

This claim is supported by the boxplot in figure 1 that shows the distribution of MPG per transmission type: cars with automatic transmission have an average of `r mean(subset(mtcars, am == 0, select = mpg)[,1])` MPG (which is equal to the intercept), compared to `r mean(subset(mtcars, am == 1, select = mpg)[,1])` MPG for cars with manual transmission.

The predictor is significant, due to its small `p-value` and a [`r confint(fit0, "am")`] confidence interval. However, with an `R-squared` of `r summary(fit0)$r.squared` it explains only `r 100 * summary(fit0)$r.squared`% of the variance. 



# Data Analysis

According to Newton's law of physics, more force is needed to move objects with a higher mass. This suggests that the weight of a car is of influence on its MPG. The next model incorporates both the transmission type and the cars' weight:

```{r}
fit1 <- lm(mpg ~ am + wt, data = mtcars)
```

```{r, results = 'asis', echo = FALSE}
xtable(summary(fit1)$coefficients, caption = "Coefficients of a multivariate model considering transmission type and weight.", digits=c(0,2,2,2,6))
```

Indeed, the weight appears to be a good predictor as indicated by the very small `p-value`. Furthermore, the model explains `r 100 * summary(fit1)$r.squared`% of the variance.

Another interesting observation is that one can see from the `p-values` in both models that holding the weight constant, transmission type appears to much have less of an impact on MPG than if weight is disregarded. This suggests that the weight of a car is far more important with respect to MPG than its transmission type. 

After trying several other variables, `qsec` was selected as third argument in the model under the reasoning is that the faster the car, the higher its fuel consumption and thus the less MPG. Table 3 lists the variance table for the three models.

```{r}
fit2 <- lm(mpg ~ am + wt + qsec, data = mtcars)
```

```{r, results = 'asis', echo = FALSE} 
xtable(anova(fit0, fit1, fit2), caption = "Analysis of variance tables for the three models.", digits=c(0,0,2,1,2,2,6))
```

Figure 3 shows some minor residual heteroskedacity and the distribution plot of the residuals of manual cars (figure 4) shows a slight shift just to the left of the mean, indicating that the residuals of the regression do not appear biased with mean close to 0. 



# Conclusion

The third model, including `am`, `wt` and `qsec` as variables explains `r 100 * summary(fit2)$r.squared`% of the variance. Additional variables did not pass the statistical significance test and/or did not contribute to variance of the model. 



\newpage

# Appendix 1: Figures

## Figure 1. Boxplot of MPG per transmission type
```{r, boxplot, echo = FALSE}
ggplot(mtcars, aes(factor(am), mpg)) + 
    geom_boxplot() +
    geom_point(size = 5, aes(colour = factor(am)), alpha = 0.6) +
    xlab("Transmission type (0 = Automatic, 1 = Manual)") + 
    ylab("Miles/(US) gallon") +
    theme_minimal(base_family = "") 
```

## Figure 2. Regression line of MPG vs. weight per transmission type
```{r, regression, echo = FALSE}
ggplot(mtcars, aes(wt, mpg)) + 
    geom_point(size = 5, aes(colour = factor(am)), alpha = 0.6) +
    geom_smooth(aes(colour = factor(am), group = am), method = "lm") +
    xlab("Weight (lb/1000)") + 
    ylab("Miles/(US) gallon") +
    theme_minimal(base_family = "") 
```

## Figure 3. Residual plot 
```{r, residualplot, echo = FALSE}
ggplot(mtcars, aes(wt, resid(fit2))) + 
    geom_hline(yintercept = 0, size = 1) +
    geom_point(size = 5, aes(colour = factor(am)), alpha = 0.6) +
    xlab("Fitted values") + 
    ylab("Residuals") +
    theme_minimal(base_family = "") 
```

## Figure 4. Residual histogram
```{r, residualhist, echo = FALSE}
ggplot(mtcars, aes(resid(fit2), fill = factor(am))) + 
    geom_density(alpha = 0.2) +
    geom_vline(yintercept = 0, size = 1, colour = "red") +
    xlab("Fitting approach") + 
    ylab("Residuals") +
    theme_minimal(base_family = "") 
```

